---
title: "Part 2: TCR"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 3
    title-block-banner: "#00A7FF"
filters:
  - flourish
css: report_assets/style.css
editor: visual
bibliography: report_assets/references.bib
---

# Single-Cell TCR Analysis Workbook with scRepertoire

## Introduction

T-cell receptor (TCR) analysis allows us to understand the adaptive immune response at single-cell resolution. Each T cell has a unique TCR sequence that defines its antigen specificity. In this workbook, we'll analyze TCR data from diseased human PBMCs using the scRepertoire package.

### What we'll learn:
- Load and process 10x TCR data
- Quantify TCR clonality and diversity  
- Visualize clonal expansion patterns
- Integrate TCR data with RNA-seq analysis

## Setup and Data Loading

```{r}
# Load required libraries
library(scRepertoire)
library(Seurat)
library(ggplot2)
library(dplyr)

# Set working directory and load RNA-seq data
scData <- qs::qread("~/work/10k-pbmc-rna-tcr-bcr/scData.qs2")

# Check the structure of our RNA-seq data
print(scData)
table(scData$seurat_clusters)
```

### Loading TCR Contigs

The 10x Genomics pipeline generates filtered contig annotations that contain TCR sequence information.

```{r}
# Load TCR contig data from 10x output
# The function will automatically detect the 10x format
tcr_data <- loadContigs("~/work/10k-pbmc-rna-tcr-bcr/", format = "10X")

# Check what we loaded
print(tcr_data)
names(tcr_data)
```

## Processing TCR Data

### Combining TCR Contigs

We need to process the raw contig data into clonal information for each cell.

```{r}
# Combine TCR data into clones
# This identifies unique TCR sequences and groups cells with identical TCRs
combined_tcr <- combineTCR(tcr_data, 
                           samples = "PBMC_diseased",
                           removeNA = FALSE,        # Keep cells without TCRs for now
                           removeMulti = FALSE,     # Keep cells with multiple TCR chains
                           filterMulti = TRUE)      # But filter to most abundant chains

# Look at the structure
print(combined_tcr)
head(combined_tcr[[1]])
```

**Key concepts:**
- **Clone**: Cells with identical TCR sequences (same V, D, J genes and CDR3 sequence)
- **cloneCall**: Different ways to define clones ("strict" uses V(D)J + nucleotide CDR3)
- **Productive vs Non-productive**: Whether the TCR sequence can produce a functional protein

### Basic TCR Statistics

```{r}
# How many cells have TCRs?
tcr_cells <- combined_tcr[[1]]
total_cells <- nrow(tcr_cells)
cells_with_tcr <- sum(!is.na(tcr_cells$CTstrict))

cat("Total cells:", total_cells, "\n")
cat("Cells with TCRs:", cells_with_tcr, "\n") 
cat("TCR detection rate:", round(cells_with_tcr/total_cells*100, 1), "%\n")
```

## Basic TCR Analysis

### 1. Clone Quantification

How many unique TCR clones do we have?

```{r}
# Plot the total number of unique clones
clonalQuant(combined_tcr, 
            cloneCall = "strict",
            scale = TRUE)  # Show as percentage
```

### 2. Clonal Abundance Distribution

How are clone sizes distributed?

```{r}
# Visualize clone abundance (frequency distribution)
clonalAbundance(combined_tcr,
                cloneCall = "strict",
                scale = FALSE)  # Show raw counts
```

**Interpretation:** Most clones are singlets (1 cell), with fewer expanded clones. This is typical of healthy immune repertoires.

### 3. Clonal Homeostasis

What proportion of the repertoire is occupied by different clone size groups?

```{r}
# Analyze clonal homeostasis (repertoire composition)
clonalHomeostasis(combined_tcr,
                  cloneCall = "strict")
```

**Interpretation:** 
- **Rare clones** (single cells): Usually most abundant
- **Hyperexpanded clones**: May indicate recent immune activation

### 4. TCR Sequence Length Distribution

```{r}
# Analyze CDR3 amino acid length distribution
clonalLength(combined_tcr,
             cloneCall = "aa",  # Use amino acid sequences
             chain = "both")    # Include both TCR chains
```

### 5. Clonal Diversity

How diverse is our TCR repertoire?

```{r}
# Calculate Shannon diversity (higher = more diverse)
clonalDiversity(combined_tcr,
                cloneCall = "strict",
                metric = "shannon")

# Compare with inverse Simpson index
clonalDiversity(combined_tcr,
                cloneCall = "strict", 
                metric = "inv.simpson")
```

**Interpretation:**
- **Shannon entropy**: Accounts for both richness (number of clones) and evenness
- **Inverse Simpson**: More sensitive to abundant clones

## Integration with RNA-seq Data

Now we'll combine our TCR analysis with the RNA-seq data to see how clonal expansion relates to cell types and states.

### Combining Expression and TCR Data

```{r}
# Add TCR information to our Seurat object
# This matches cells by barcode between datasets
scData <- combineExpression(combined_tcr, 
                           scData,
                           cloneCall = "strict",
                           proportion = FALSE,  # Use raw clone counts
                           cloneSize = c(Single = 1, Small = 5, Medium = 20, Large = 100, Hyperexpanded = 500))

# Check what was added
colnames(scData@meta.data)[grep("clone|CTstrict", colnames(scData@meta.data))]

# How many cells got TCR information?
table(is.na(scData$CTstrict))
```

### TCR Metrics by Cell Cluster

```{r}
# Visualize clone sizes across clusters
clonalOccupy(scData, 
             x.axis = "seurat_clusters")
```

**Questions to consider:**
- Which clusters have the most expanded clones?
- Are certain cell types more clonally expanded?

### Advanced Visualization: Clonal Overlay

This shows where expanded clones are located in UMAP space:

```{r}
# Overlay clonal expansion on UMAP
# Shows cells with clones above frequency threshold
clonalOverlay(scData,
              reduction = "umap",
              cutpoint = 5,        # Show clones with ≥5 cells
              bins = 10)           # Smoothing for density contours
```

**Interpretation:** Expanded clones may cluster in specific regions, suggesting shared functions or activation states.

## Gene Expression in Expanded Clones

### Finding Clonally Expanded Cells

```{r}
# Identify cells in expanded clones (>= 5 cells per clone)
expanded_clones <- scData@meta.data$clonalFrequency >= 5
table(expanded_clones, useNA = "ifany")

# Add this as metadata
scData$expanded_clone <- ifelse(expanded_clones & !is.na(expanded_clones), 
                               "Expanded", "Single/Small")
table(scData$expanded_clone, useNA = "ifany")
```

### Visualization on UMAP

```{r}
# Plot UMAP colored by clonal expansion
DimPlot(scData, 
        group.by = "expanded_clone",
        reduction = "umap") +
  ggtitle("Clonally Expanded vs Single/Small Clones")

# Also show by clone size categories
DimPlot(scData,
        group.by = "cloneSize", 
        reduction = "umap") +
  ggtitle("Clone Size Distribution")
```

## Summary Statistics

```{r}
# Create a summary table
expansion_summary <- scData@meta.data %>%
  filter(!is.na(CTstrict)) %>%  # Only cells with TCRs
  group_by(seurat_clusters) %>%
  summarise(
    total_tcr_cells = n(),
    n_unique_clones = length(unique(CTstrict)),
    n_expanded_clones = sum(clonalFrequency >= 5, na.rm = TRUE),
    pct_expanded = round(n_expanded_clones/total_tcr_cells*100, 1),
    .groups = 'drop'
  ) %>%
  arrange(desc(pct_expanded))

print(expansion_summary)
```

## Key Takeaways

1. **TCR Detection**: Not all cells will have detected TCRs (~30-70% typical for PBMCs)

2. **Clonal Structure**: Most TCRs are unique (singlets), with few expanded clones

3. **Disease Context**: In diseased samples, you might see:
   - More clonal expansion than healthy controls
   - Specific cell types with higher expansion rates
   - Potential oligoclonal responses

4. **Integration Value**: Combining TCR and RNA-seq data reveals:
   - Which cell types undergo clonal expansion
   - Whether expanded clones have distinct gene expression
   - Spatial organization of clones in transcriptional space

## Next Steps

For more advanced analysis, consider:
- Comparing with healthy controls
- Gene set enrichment in expanded clones  
- TCR motif analysis
- Clonal tracking across conditions/timepoints
- BCR analysis if B cell data available

## Troubleshooting

**Low TCR detection rate?**
- Check if 5' or 3' chemistry was used (5' captures TCRs better)
- Verify cell viability and RNA quality
- Consider TCR enrichment protocols

**No expanded clones?**
- Normal in some healthy samples
- May need deeper sequencing
- Consider different clone size thresholds

**Integration issues?**
- Check barcode formats match between datasets
- Verify cell filtering was applied consistently
- Consider batch effects if multiple samples

